Practical Hilbert space approximate Bayesian Gaussian processes for probabilistic programming

### 2022-03-14 Paper/Manuscript_Stats&Comp/manuscript_STCO.tex
latest/submitted version of revised munuscript and rebuttal

% There is a clear relation between the number of basis functions $m$ and the boundary factor $c$ with the length-scale $\ell$ of the approximated function. In this section we discuss about the existing theoretical evidence that the minimum number of basis functions $m$ needed for an accurate approximation is linearly proportional to the the boundary factor $c$ and inversely proportional to the length-scale $l$.
% In the next Section \ref{subsec_empiric_relations} we conduct an empirical analysis through simulations to prove this expected linear proportionality between $m$, $c$ and $l$ and compute the specific coefficients of proportionality for every kernel.

% {\color{blue}Aki: We can also look at the mean zero-level upcrossings, which for QE is
% proportional to 1/l which gives as that part, and the effect of c should also
% be quite linear when c is not very close to 1 and m is not very small as it
% linearly makes the interval shorter so that the there are less zero-level
% upcrossings available in the basis functions. You can visualize this using
% {\small \url{https://avehtari.github.io/casestudies/Motorcycle/motorcycle_gpcourse.html}} and
% look how the plot of 6 basis functions change when you set c=1, c=1.5, c=2.25
% (ie $1.5^2$). Instead of looking at mean zero-level crossings, we can look at the
% spectral densities and when they go to zero. This is illustrated in the same
% notebook after that figure. E.g. with length-scale l=1,l=1/2, l=1/4, l=1/8 to
% explain 99\% of the prior variance we need 4, 8, 16, and 32 basis functions
% (linear in l) and to explain 80\% of the prior variance we need 2, 4, 8, 16
% basis functions. These spectral densities don't depend on c. When c increaes
% the basis functions are truncated and for example going from c=1.5 to c=2.25,
% basis function 9 has the flexibility basis function 6 had, so 1.5 more basis
% functions are needed. So we have proportionality c/l. }

% {\color{red} Gabi: The spectral densities do depend on $c$, see equations (\ref{eq_approxf}) and (\ref{eq_eigenvalue}).}

@@

% From the empirical observations the hypothesis of linear proportionality between $m$, $\frac{l}{S}$ (lenghscale $l$ relative to the half range of the input domain $S$) and $c$ has been proved. Following we show the numerical functions governing these relationships for every kernel.

@@

Apart from providing a powerful diagnostic tool in determining if the approxiamtion is sufficiently accurate, the equations in the previous Section~\ref{sec_num_equations} also provide the optimal values for $m$ (the minimum $m$ required for an accurate approximation) and $c$ (the minimum $c$ that allows for the minimum $m$) that van be used to minimize the computational cost in repeated computations (e.g., in cross-validation and simulation based calibration)
%If the initial lengthscale is not far from the optimal, and larger than minimal $c$ and $m$ are used in the initial HSGP model, it is likely that the initial results are already sufficiently accurate. 
%The iterative procedure, briefly commented above and detailed in next Section~\ref{sec_user_guide}, works as safety approach and also allows finding a smaller sufficient $m$ which can be useful if the user wants to optimize the computational speed.
%Using the optimum values for $m$ and $c$ that allow for minimum computational requirements can be essential for repeated fitting (e.g., in cross-validation or simulation based calibration), uni-dimensional functions with very large data sets, and
This is even more useful in multi-dimensional cases ($2 \leq D \leq 4$), where knowing the smallest useful value of $m$ for each dimension has even bigger effect on the total computational cost.

@@

Based on the above proposed diagnostics, we obtain a simple, iterative step-by-step procedure that users can follow in order to obtain an accurate HSGP approximation. The procedure is split into two phases, Phase~A and B, which have to be completed consecutively.
%In the following, we will refer to the Equations from Section XXX for the double-exponential kernel, but the same procedure works for other kernels by using the corresponding Equations provided in the same Section.

@@

% \subsubsection{Criticism of the diagnostic} \label{sec_diagnostic_criticism}

% The basis of our diagnostic tool are the following:
% % 
% \begin{enumerate}
% \item observing a length-scale estimate $\hat{\ell}$ smaller than recommended minimum length-scale determined by $m$ and $c$ indicates a potential problem with the approximation,
% \item observing a $\hat{\ell}$ larger than the recommended minimum length scale implies that the approximation is accurate.
% \end{enumerate}

% The later statement, however, cannot be assumed to be true in general, that is, cannot be generalized to all possible cases. ({\color{red} TODO: A major revision from a reviewer was related to that, i.e. to be critical with the diagnostic and discuss about possible cases where it could fail. I don't know if we could discuss a bit about this. I remember Michael commented an example of a potential case where that diagnostic statement could not be true.})

% For that reason, to ultimately diagnose whether the approximation is accurate, in addition to observe whether $\hat{\ell}$ exceeds the recommended minimum length scale, we recommend ensuring the measures of predictive accuracy reach an stabilization too.

@@

%If much bigger $m$ than necessary is used (i.e., too much large $m$ relative to the wigglyness of the function to be learned), there can be computational challenges. In Figure~\ref{fig7_lscale_comparison}, we see that increasing $m$ much beyond what is needed, is leading to underestimating the lengthscale. Examining the convergence diagnostics revealed problems in exploring larger length-scales, which is likely due to the posterior having a difficult funnel shape \citep[see, e.g.,][]{Betancourt+Girolami:2019}. A largely underestimation of the length-scale leads to overestimating the underlying functions (see Figure~\ref{fig8_overestimated_posterios} as an illustration of the overestimating effect as a function of $m$ in a moderate smooth dataset). Increasing the boundary factor $c$ (more than that recommended in equations~\eqref{sec_num_equations}) reduces the funnellness of the posterior and the sampling problems in exploring larger length-scales mostly disappear and so underestimating the length-scale.
%
%Following our recommended diagnostic tool is a safety procedure to avoid using too much large $m$ that needed, as it allows us to select optimal values for $m$ and $c$ in each iteration.
%
%\begin{figure}
%\centering
%\includegraphics[width=0.5\textwidth, trim = 0mm 0mm 0mm 0mm, clip]{fig8_overestimated_posterios.png}
%\caption{Overestimation of the data points when using too many number of basis functions relative to the wigglyness of the data points. The dashed red line represents the true underlying function.}
%  \label{fig8_overestimated_posterios}
%\end{figure}

Table~\ref{tab1_diagnostic} shows the iterative steps of applying the diagnostic procedure explained in Section~\ref{sec_user_guide} over some of the data sets also used in the analysis in Figure~\ref{fig7_lscale_comparison} left. It is clearly visible that by following our recommendations, an optimum solution with minimum computational requirements is achieved in these cases. Figure~\ref{fig10_diagnostic_lscale_comparison} graphically compares the exact GP length-scale and the estimated HSGP length-scale in every iteration and data set. Between two and four iterations, depending on wigglyness of the function to be learned and the distance between the initial guess of the length scale and the true length scale, are sufficient to reach the optimal values of $m$ and $c$. 
%Furthermore, as can be seen in Figure~\ref{fig10_diagnostic_lscale_comparison}, following our diagnostic recommendations avoids the use of a too large number of basis functions relative to the true length scale that would cause overestimating the underlying function and bias the posterior towards smaller length scale values.

@@

\begin{table*}
\centering
\setlength{\tabcolsep}{5pt}
\begin{tabular}{ c c c c c c c c c c | c c}
\arrayrulecolor{gray}\hline \\[-3mm]
$\ell_{\text{GP}}$ & iter. & $\ell$ & $c$ & $m$ & $\hat{\ell}$ & $\hat{\ell} + 0.01 \geq \ell$ & RMSE & $R^2$ & ELPD & RMSE* & \\ 
%\arrayrulecolor{lightgray}\hline \\[-1mm]
%\multirow{5}{*}{ 0.05} 
% & 1 & 0.50 & 1.60 & 6 & 0.127 & FALSE & 0.800 & 0.264 & -0.980 & 0.748 & \multirow{5}{*}{ \includegraphics[scale=0.20, trim = 0mm 10mm 0mm 5mm, clip]{fig9_diagnostic_1.png}}\\
% & 2 & 0.127 & 1.20 & 17 & 0.061 & FALSE & 0.427 & 0.779 & -0.396 & 0.330 & \\
% & 3 & 0.061 & 1.20 & 35 & 0.059 & TRUE & 0.264 & 0.908 & 0.008 & 0.063 & \\
% & 4 & 0.053 & 1.20 & 40 & 0.059 & TRUE & 0.254 & 0.915 & 0.088 & 0.029 & \\
% \\[1mm]
\arrayrulecolor{lightgray}\hline \\[-1mm]
\multirow{5}{*}{ 0.08} 
 & 1 & 0.50 & 1.60 & 6 & 0.17 & FALSE & 0.45 & 0.61 & -0.40 & 0.36 & \multirow{5}{*}{ \includegraphics[scale=0.20, trim = 0mm 10mm 0mm 5mm, clip]{fig9_diagnostic_2.png}}\\
 & 2 & 0.17 & 1.20 & 13 & 0.07 & FALSE & 0.36 & 0.74 & -0.18 & 0.24 & \\
 & 3 & 0.07 & 1.20 & 31 & 0.08 & TRUE & 0.24 & 0.87 & 0.20 & 0.01 & \\
 & 4 & 0.06 & 1.20 & 36 & 0.08 & TRUE & 0.24 & 0.87 & 0.21 & 0.01 & \\
 \\[1mm]
 \arrayrulecolor{lightgray}\hline \\[-1mm]
\multirow{3}{*}{ 0.13} 
 & 1 & 0.50 & 1.60 & 6 & 0.25 & FALSE & 0.46 & 0.63 & -0.49 & 0.33 & \multirow{5}{*}{ \includegraphics[scale=0.20, trim = 0mm 10mm 0mm 5mm, clip]{fig9_diagnostic_3.png}}\\
 & 2 & 0.25 & 1.20 & 9 & 0.15 & FALSE & 0.35 & 0.78 & -0.14 & 0.15 & \\
 & 3 & 0.15 & 1.20 & 15 & 0.15 & TRUE & 0.32 & 0.81 & -0.07 & 0.08 & \\
 & 4 & 0.11 & 1.20 & 20 & 0.12 & TRUE & 0.30 & 0.83 & -0.01 & 0.01 & \\
 \\[1mm]
\arrayrulecolor{lightgray}\hline \\[-1mm]
\multirow{3}{*}{ 0.25} 
 & 1 & 0.50 & 1.60 & 6 & 0.24 & FALSE & 0.37 & 0.40 & -0.15 & 0.21 & \multirow{5}{*}{ \includegraphics[scale=0.20, trim = 0mm 10mm 0mm 5mm, clip]{fig9_diagnostic_4.png}}\\
 & 2 & 0.24 & 1.20 & 9 & 0.26 & TRUE & 0.30 & 0.60 & -0.09 & 0.02 & \\
 & 3 & 0.25 & 1.20 & 14 & 0.26 & TRUE & 0.29 & 0.62 & -0.05 & 0.01 & \\
 \\
 \\[1mm]
\arrayrulecolor{lightgray}\hline \\[-1mm]
\multirow{3}{*}{ 0.54} 
 & 1 & 0.50 & 1.60 & 6 & 0.38 & FALSE & 0.27 & 0.86 & 0.03 & 0.01 & \multirow{5}{*}{ \includegraphics[scale=0.20, trim = 0mm 10mm 0mm 3mm, clip]{fig9_diagnostic_5.png}}\\
 & 2 & 0.38 & 1.20 & 6 & 0.47 & TRUE & 0.28 & 0.86 & 0.03 & 0.05 & \\
 & 3 & 0.24 & 1.50 & 11 & 0.53 & TRUE & 0.27 & 0.86 & 0.04 & 0.01 & \\
 \\
 \\[1mm]
\arrayrulecolor{lightgray}\hline \\[-1mm]
\multirow{3}{*}{ 0.80} 
 & 1 & 1.00 & 3.20 & 6 & 0.52 & FALSE & 0.29 & 0.92 & 0.13 & 0.03 & \multirow{5}{*}{ \includegraphics[scale=0.20, trim = 0mm 10mm 0mm 3mm, clip]{fig9_diagnostic_6.png}}\\
 & 2 & 0.52 & 1.65 & 6 & 0.98 & TRUE & 0.29 & 0.92 & 0.11 & 0.03 & \\
 & 3 & 0.50 & 3.12 & 11 & 0.77 & TRUE & 0.28 & 0.92 & 0.11 & 0.01 & \\
 \\
 \\[1mm]
\arrayrulecolor{lightgray}\hline \\[-1mm]
\multirow{2}{*}{ 1.40} 
 & 1 & 1.00 & 3.20 & 6 & 1.02 & TRUE & 0.28 & 0.56 & 0.14 & 0.01 & \multirow{5}{*}{ \includegraphics[scale=0.20, trim = 0mm 10mm 0mm 3mm, clip]{fig9_diagnostic_7.png}}\\
 & 2 & 0.52 & 3.27 & 11 & 1.23 & TRUE & 0.28 & 0.57 & 0.14 & 0.00 & \\
 \\
 \\
 \\[1mm]
 \arrayrulecolor{lightgray}\hline \\[-1mm]
\multirow{2}{*}{ 2.94} 
 & 1 & 0.50 & 1.60 & 6 & 1.19 & TRUE & 0.34 & 0.45 & -0.10 & 0.06 & \multirow{5}{*}{ \includegraphics[scale=0.20, trim = 0mm 10mm 0mm 3mm, clip]{fig9_diagnostic_8.png}}\\
 & 2 & 0.61 & 3.81 & 11 & 2.58 & TRUE & 0.34 & 0.45 & -0.07 & 0.01 & \\
 \\
 \\
 \\[1mm]
\arrayrulecolor{gray}\hline
\end{tabular}
\caption{diagnostic on various data sets with varying smoothness. RMSE* refers to the root mean squared error computed with the exact GP model as the reference.}
  \label{tab1_diagnostic}
\end{table*}

\begin{figure}
\centering
\subfigure{\includegraphics[width=0.5\textwidth, trim = 0mm 0mm 0mm 0mm, clip]{fig10_diagnostic_lscale_comparison.png}}
\caption{Comparison of the true and estimated length-scales in every iteration for the different data sets varying the length-scale.}
  \label{fig10_diagnostic_lscale_comparison}
\end{figure}

@@

The HSGP model is computationally superior in $1$D and $2$D even for highly wiggly functions, except when the number of data points is so small ($n \lesssim 300$) that exact GPs are already reasonably fast themselves. However, the computation time of the HSGP model increases rapidly with the number of input dimensions ($D$) since the number of multivariate basis functions $m^*=m_1\times \cdots \times m_D$ in the approximation increases exponentially with $D$ (see eq.~(\ref{eq_m_multi})). Yet, the HSGP method can still be computationally faster than the exact GP for larger datasets due the latter's cubic scaling in $n$.

In our experiments of multivariate problems (see Section~\ref{sec_computation_nD}), the computation time for the HSGP model was faster than for the exact GP for most of the non-linear $2D$ functions or moderate-to-large sized $3$D datasets ($n \gtrsim 1000$), even for highly wiggly 3D functions (e.g., $\ell_1, \ell_2, \ell_3 \approx 0.1$).

{\color{blue} For small sized datasets ($n\lesssim 1000$), HSGPs are likely to be slower than exact GPs already for moderate to highly wiggly $3$D functions (e.g., $\ell_1 \lesssim 0.1$, and $\ell_2, \ell_3 \gtrsim 0.3$) and for overall smooth $4$D functions (e.g., $\ell_1 \lesssim 0.1$, and $\ell_2, \ell_3, \ell_4 \gtrsim 0.4$).}

%In all of the investigated cases, choosing the optimal booundary factor $c$ reduces the required number of basis functions $m$ noticeably (see Figures~A.3, B.3-left and C.2 in the online material) and therefore  reduces computational time drastically in particular in multidimensional cases.

As it has been shown in case study III (Section~\ref{sec_caseIII}), the proposed diagnostic tool can be very useful for multivariate problems as it allows one to reduce $m^*$ to the minimum sufficient value, reducing computational time drastically, and still getting an accurate approximation. {\color{blue} For example, assuming a squared exponential covariance function, choosing the optimal value for $c$ allows one to use few basis functions in every single dimension (i.e., $m \lesssim 10$ for $\ell \gtrsim 0.3$; $20 \gtrsim m \gtrsim 10$ for $0.3 \gtrsim \ell \gtrsim 0.1$; and $m \gtrsim 20$ for $\ell \lesssim 0.1$), which, from results presented in Figure~\ref{fig_time_nD_log}, implies that the HSGP model can be, in general terms, useful for highly wiggly $3$D functions and smooth $4$D functions.}

@@

In our experiments, we observed roughly similar or even worse scaling behavior for the thin plate splines in comparison to HSGPs, with serious challenges with computation time when doing inference with spline models for $D=3$ and more than 10 basis functions per dimension, or even for $D=2$ and more than 40 basis functions per dimension (see Figures~B.3-right and C.3 in the online supplement).

@@

In this experiment, we analyze a synthetic dataset with $n = 250$ observations, where the true data generating process is a Gaussian process with additive noise. The data points are simulated from the model $y_i = f(x_i) + \epsilon_i$, where $f$ is a sample from a Gaussian process $f(x) \sim \GP(0, k(x, x', \theta))$ with a Mat{\'e}rn ($\nu$=3/2) covariance function $k$ with marginal variance $\alpha=1$ and length-scale $\ell=0.2$ at inputs values $\bm{x}=(x_1,x_2,\dots,x_n)$ with $x_i \in [-1,1]$. $\epsilon_i$ is additive Gaussian noise with standard deviation $\sigma=0.2$. %We split the dataset into three parts: $155$ data points are used for fitting the model (training set), $45$ data ponts are used for the interpolation test set, and the remaining $50$ data points are used for the extrapolation test set.
%
%The exact GP model for fitting and predicting this simulated dataset $\bm{y}$ can be written as follows,
%%
%\begin{align*}\label{eq_latentgp_simudata1}
%\begin{split}
%\bm{y} &= \bm{f} + \bm{\epsilon} \\
%\bm{\epsilon} &\sim \Normal(\bm{0}, \sigma^2 \bm{I}) \\
%f(x) &\sim \GP(0, k(x, x', \theta)),
%\end{split}
%\end{align*}
%%
%where $\bm{f}=\{f(x_i)\}_{i=1}^n$ represents the underlying function at the input values $x_i$, and $\bm{\epsilon}$ is the Gaussian noise term with variance $\sigma^2$, with $\bm{I}$ representing the identity matrix. The function $f:{\rm I\!R} \to {\rm I\!R}$ is a GP prior with the Mat{\'e}rn ($\nu$=3/2) covariance function $k$, which depends on the inputs $\bm{x}$ and hyperparameters $\theta=\{\alpha,\ell\}$. The hyperparameters $\alpha$ and $\ell$ represent the marginal variance and length-scale, respectively, of the GP process. Saying that the function $f(\cdot)$ follows a GP model is equivalent to say that $\bm{f}$ is multivariate Gaussian distributed with covariance matrix $\bm{K}$, where $K_{ij}=k(x_i,x_j,\theta)$, with $i,j=1,2,\dots,n$.
% 
%A more computationally efficient formulation of a GP model with Gaussian likelihood, and for probabilistic inference using MCMC sampling methods, would be its marginalized form,
%%
%\begin{equation*}\label{eq_marginalizedgp_simudata1}
%\bm{y} \sim \Normal(\bm{0},\bm{K}+ \sigma^2 \bm{I} ),
%\end{equation*}
%%
%where the function values $\bm{f}$ have been integrated out, yielding a lower-dimensional parameter space over which to do inference, reducing the time of computation and improving the sampling and the effective number of samples.

In the HSGP model, the latent function values $f(x)$ are approximated as in eq.~(\ref{eq_approxf}), with the Mat{\'e}rn ($\nu$=3/2) spectral density $S$ as in eq.~(\ref{eq_specdens_32}), and eigenvalues $\lambda_j$ and eigenfunctions $\phi_j$ as in equations (\ref{eq_eigenvalue}) and (\ref{eq_eigenfunction}), respectively.  %In order to do model comparison, in addition to the exact GP model and HSGP model, a spline-based model is also fitted using the thin plate regression spline approach by \citet{wood2003thin} and implemented in the R-package \textit{mgcv} \citep{wood2011mgcv}. A Bayesian approach is used to fit this spline model using the R-package \textit{brms} \citep{burkner2017brms}.

@@

\begin{table}
\centering
\setlength{\tabcolsep}{4pt}
\begin{tabular}{cccccccc}
\arrayrulecolor{gray}\hline \\[-3mm]
iter. & $\ell$ & $c$ & $m$ & $\hat{\ell}$ & $\hat{\ell} + 0.01 \geq \ell$ & RMSE & RMSE* \\ 
\arrayrulecolor{lightgray}\hline \\[-1mm]
 1 & 0.50 & 2.25 & 16 & 0.12 & FALSE & 0.26 & 0.17 \\
 2 & 0.12 & 1.20 & 35 & 0.32 & TRUE & 0.18 & 0.01 \\
 3 & 0.02 & 1.20 & 40 & 0.28 & TRUE & 0.18 & 0.01 \\[1mm]
\arrayrulecolor{gray}\hline
\end{tabular}
\caption{diagnostic on various data sets with varying smoothness. RMSE* refers to the root mean squared error computed with the exact GP model as the reference.}
  \label{tab_caseI}
\end{table}

Figure~\ref{fig11_Posteriors_exI} shows the posteriors predictive distributions of the exact GP and the HSGP models, the later using the parameter values as the third iterative step, $c=1.2$ ($L=c\cdot 1= 1.2$; see eq.~\eqref{eq_boundary}) and $m=40$ basis functions. In order to do model comparison a spline-based model with 60 knots is also fitted and plotted jointly with the GP and HSGP models. The spline-based model is fitted using the thin plate regression spline approach by \citet{wood2003thin} and implemented in the R-package \textit{mgcv} \citep{wood2011mgcv}. A Bayesian approach is used to fit this spline model using the R-package \textit{brms} \citep{burkner2017brms}. The sample observations are plotted as circles and the out-of-sample or test data are plotted as crosses. The posteriors of the three models, exact GP, HSGP and spline, are similar in the interpolation regions of the input space. However, when extrapolating the spline model solution clearly differs from the exact GP and HSGP models as well as the actual observations.

\begin{figure}
\centering
\includegraphics[width=0.5\textwidth]{fig11_Posteriors_exI.png}
\caption{Posterior predictive means of the proposed HSGP model, the exact GP model, and the spline model. 95\% credible intervals are plotted as dashed lines.}
  \label{fig11_Posteriors_exI}
\end{figure}

>>>>

@@

@@

Apart from providing a powerful diagnostic tool in determining if the approxiamtion is sufficiently accurate, the equations in the previous Section~\ref{sec_num_equations} also provide the optimal values for $m$ (the minimum $m$ required for an accurate approximation) and $c$ (the minimum $c$ that allows for the minimum $m$) that van be used to minimize the computational cost in repeated computations (e.g., in cross-validation and simulation based calibration).
This is even more useful in multi-dimensional cases ($2 \leq D \leq 4$), where knowing the smallest useful value of $m$ for each dimension has even bigger effect on the total computational cost.

@@

Based on the above proposed diagnostics, we obtain a simple, iterative step-by-step procedure that users can follow in order to obtain an accurate HSGP approximation. The procedure is split into two phases, Phase~A and B, which have to be completed consecutively.

@@

@@

Table~\ref{tab1_diagnostic} shows the iterative steps of applying the diagnostic procedure explained in Section~\ref{sec_user_guide} over some of the data sets also used in the analysis in Figure~\ref{fig7_lscale_comparison} left. It is clearly visible that by following our recommendations, an optimum solution with minimum computational requirements is achieved in these cases. Figure~\ref{fig8_diagnostic_lscale_comparison} graphically compares the exact GP length-scale and the estimated HSGP length-scale in every iteration and data set. Between two and four iterations, depending on wigglyness of the function to be learned and the distance between the initial guess of the length scale and the true length scale, are sufficient to reach the optimal values of $m$ and $c$.

@@

\begin{table*}
\centering
\setlength{\tabcolsep}{5pt}
\begin{tabular}{ c c c c c c c c c c | c c}
\arrayrulecolor{gray}\hline \\[-3mm]
$\ell_{\text{GP}}$ & iter. & $\ell$ & $c$ & $m$ & $\hat{\ell}$ & $\hat{\ell} + 0.01 \geq \ell$ & RMSE & $R^2$ & ELPD & RMSE* & \\ 
\arrayrulecolor{lightgray}\hline \\[-1mm]
\multirow{5}{*}{ 0.08} 
 & 1 & 0.50 & 1.60 & 6 & 0.17 & FALSE & 0.45 & 0.61 & -0.40 & 0.36 & \multirow{5}{*}{ \includegraphics[scale=0.20, trim = 0mm 10mm 0mm 5mm, clip]{tab1_diagnostic_2.png}}\\
 & 2 & 0.17 & 1.20 & 13 & 0.07 & FALSE & 0.36 & 0.74 & -0.18 & 0.24 & \\
 & 3 & 0.07 & 1.20 & 31 & 0.08 & TRUE & 0.24 & 0.87 & 0.20 & 0.01 & \\
 & 4 & 0.06 & 1.20 & 36 & 0.08 & TRUE & 0.24 & 0.87 & 0.21 & 0.01 & \\
 \\[1mm]
 \arrayrulecolor{lightgray}\hline \\[-1mm]
\multirow{3}{*}{ 0.13} 
 & 1 & 0.50 & 1.60 & 6 & 0.25 & FALSE & 0.46 & 0.63 & -0.49 & 0.33 & \multirow{5}{*}{ \includegraphics[scale=0.20, trim = 0mm 10mm 0mm 5mm, clip]{tab1_diagnostic_3.png}}\\
 & 2 & 0.25 & 1.20 & 9 & 0.15 & FALSE & 0.35 & 0.78 & -0.14 & 0.15 & \\
 & 3 & 0.15 & 1.20 & 15 & 0.15 & TRUE & 0.32 & 0.81 & -0.07 & 0.08 & \\
 & 4 & 0.11 & 1.20 & 20 & 0.12 & TRUE & 0.30 & 0.83 & -0.01 & 0.01 & \\
 \\[1mm]
\arrayrulecolor{lightgray}\hline \\[-1mm]
\multirow{3}{*}{ 0.25} 
 & 1 & 0.50 & 1.60 & 6 & 0.24 & FALSE & 0.37 & 0.40 & -0.15 & 0.21 & \multirow{5}{*}{ \includegraphics[scale=0.20, trim = 0mm 10mm 0mm 5mm, clip]{tab1_diagnostic_4.png}}\\
 & 2 & 0.24 & 1.20 & 9 & 0.26 & TRUE & 0.30 & 0.60 & -0.09 & 0.02 & \\
 & 3 & 0.25 & 1.20 & 14 & 0.26 & TRUE & 0.29 & 0.62 & -0.05 & 0.01 & \\
 \\
 \\[1mm]
\arrayrulecolor{lightgray}\hline \\[-1mm]
\multirow{3}{*}{ 0.54} 
 & 1 & 0.50 & 1.60 & 6 & 0.38 & FALSE & 0.27 & 0.86 & 0.03 & 0.01 & \multirow{5}{*}{ \includegraphics[scale=0.20, trim = 0mm 10mm 0mm 3mm, clip]{tab1_diagnostic_5.png}}\\
 & 2 & 0.38 & 1.20 & 6 & 0.47 & TRUE & 0.28 & 0.86 & 0.03 & 0.05 & \\
 & 3 & 0.24 & 1.50 & 11 & 0.53 & TRUE & 0.27 & 0.86 & 0.04 & 0.01 & \\
 \\
 \\[1mm]
\arrayrulecolor{lightgray}\hline \\[-1mm]
\multirow{3}{*}{ 0.80} 
 & 1 & 1.00 & 3.20 & 6 & 0.52 & FALSE & 0.29 & 0.92 & 0.13 & 0.03 & \multirow{5}{*}{ \includegraphics[scale=0.20, trim = 0mm 10mm 0mm 3mm, clip]{tab1_diagnostic_6.png}}\\
 & 2 & 0.52 & 1.65 & 6 & 0.98 & TRUE & 0.29 & 0.92 & 0.11 & 0.03 & \\
 & 3 & 0.50 & 3.12 & 11 & 0.77 & TRUE & 0.28 & 0.92 & 0.11 & 0.01 & \\
 \\
 \\[1mm]
\arrayrulecolor{lightgray}\hline \\[-1mm]
\multirow{2}{*}{ 1.40} 
 & 1 & 1.00 & 3.20 & 6 & 1.02 & TRUE & 0.28 & 0.56 & 0.14 & 0.01 & \multirow{5}{*}{ \includegraphics[scale=0.20, trim = 0mm 10mm 0mm 3mm, clip]{tab1_diagnostic_7.png}}\\
 & 2 & 0.52 & 3.27 & 11 & 1.23 & TRUE & 0.28 & 0.57 & 0.14 & 0.00 & \\
 \\
 \\
 \\[1mm]
 \arrayrulecolor{lightgray}\hline \\[-1mm]
\multirow{2}{*}{ 2.94} 
 & 1 & 0.50 & 1.60 & 6 & 1.19 & TRUE & 0.34 & 0.45 & -0.10 & 0.06 & \multirow{5}{*}{ \includegraphics[scale=0.20, trim = 0mm 10mm 0mm 3mm, clip]{tab1_diagnostic_8.png}}\\
 & 2 & 0.61 & 3.81 & 11 & 2.58 & TRUE & 0.34 & 0.45 & -0.07 & 0.01 & \\
 \\
 \\
 \\[1mm]
\arrayrulecolor{gray}\hline
\end{tabular}
\caption{Diagnostic on various data sets with varying smoothness. $\ell_{GP}$ refers to the exact GP length-scale estimate. $\ell$ refers to the minimum valid length-scale that can be accurately estimated determined by $m$ and $c$ as per Section \ref{sec_num_equations}, except for the first iteration which is the initial guess of the length-scale. $\hat{\ell}$ refers to the HSGP length-scale estimate. RMSE* refers to the root mean squared error computed with the exact GP model as the reference.}
  \label{tab1_diagnostic}
\end{table*}

\begin{figure}
\centering
\subfigure{\includegraphics[width=0.5\textwidth, trim = 0mm 0mm 0mm 0mm, clip]{fig8_diagnostic_lscale_comparison.png}}
\caption{Comparison of the true and estimated length-scales in every iteration for the different data sets varying the length-scale.}
  \label{fig8_diagnostic_lscale_comparison}
\end{figure}

@@

The HSGP model is computationally superior in $1$D and $2$D even for highly wiggly functions, except when the number of data points is so small ($n \lesssim 300$, i.e., $n$ smaller than some value around 300) that exact GPs are already reasonably fast themselves. However, the computation time of the HSGP model increases rapidly with the number of input dimensions ($D$) since the number of multivariate basis functions $m^*=m_1\times \cdots \times m_D$ in the approximation increases exponentially with $D$ (see eq.~(\ref{eq_m_multi})). Yet, the HSGP method can still be computationally faster than the exact GP for larger datasets due the latter's cubic scaling in $n$.

In our experiments of multivariate problems (see Section~\ref{sec_computation_nD}), the computation time for the HSGP model was faster than for the exact GP for most of the non-linear $2D$ functions or moderate-to-large sized $3$D datasets ($n \gtrsim 1000$, i.e., $n$ greater than some value around 1000), even for highly wiggly 3D functions (e.g., $\ell_1, \ell_2, \ell_3 \approx 0.1$).

For small sized datasets ($n\lesssim 1000$), HSGPs are likely to be slower than exact GPs already for highly to moderated wiggly $3$D functions (e.g., $\ell_1 \lesssim 0.1$, and $\ell_2, \ell_3 \gtrsim 0.3$) and for overall smooth $4$D functions (e.g., $\ell_1 \lesssim 0.1$, and $\ell_2, \ell_3, \ell_4 \gtrsim 0.4$).

As it has been shown in case study III (Section~\ref{sec_caseIII}), the proposed diagnostic tool can be very useful for multivariate problems as it allows one to reduce $m^*$ to the minimum sufficient value, reducing computational time drastically, and still getting an accurate approximation. For example, assuming a squared exponential covariance function, choosing the optimal value for $c$ allows one to use few basis functions in every single dimension (i.e., $m \lesssim 10$ for $\ell \gtrsim 0.3$; $20 \gtrsim m \gtrsim 10$ for $0.3 \gtrsim \ell \gtrsim 0.1$; and $m \gtrsim 20$ for $\ell \lesssim 0.1$), which, from results presented in Figure~\ref{fig15_time_nD_log}, implies that the HSGP model can be, in general terms, useful for highly wiggly $3$D functions and smooth $4$D functions.

@@

In our experiments, we observed roughly similar or even worse scaling behavior for the thin plate splines in comparison to HSGPs, with serious challenges with computation time when doing inference with spline models for $D=3$ and more than 10 basis functions per dimension, or even for $D=2$ and more than 40 basis functions per dimension (see Figures~B.3 right and C.3 in the online supplement).

@@

In this experiment, we analyze a synthetic dataset with $n = 250$ observations, where the true data generating process is a Gaussian process with additive noise. The data points are simulated from the model $y_i = f(x_i) + \epsilon_i$, where $f$ is a sample from a Gaussian process $f(x) \sim \GP(0, k(x, x', \theta))$ with a Mat{\'e}rn ($\nu$=3/2) covariance function $k$ with marginal variance $\alpha=1$ and length-scale $\ell=0.2$ at inputs values $\bm{x}=(x_1,x_2,\dots,x_n)$ with $x_i \in [-1,1]$. $\epsilon_i$ is additive Gaussian noise with standard deviation $\sigma=0.2$.

In the HSGP model, the latent function values $f(x)$ are approximated as in eq.~(\ref{eq_approxf}), with the Mat{\'e}rn ($\nu$=3/2) spectral density $S$ as in eq.~(\ref{eq_specdens_32}), and eigenvalues $\lambda_j$ and eigenfunctions $\phi_j$ as in equations (\ref{eq_eigenvalue}) and (\ref{eq_eigenfunction}), respectively.

@@

\begin{table}
\centering
\setlength{\tabcolsep}{4pt}
\begin{tabular}{cccccccc}
\arrayrulecolor{gray}\hline \\[-3mm]
iter. & $\ell$ & $c$ & $m$ & $\hat{\ell}$ & $\hat{\ell} + 0.01 \geq \ell$ & RMSE & RMSE* \\ 
\arrayrulecolor{lightgray}\hline \\[-1mm]
 1 & 0.50 & 2.25 & 16 & 0.12 & FALSE & 0.26 & 0.17 \\
 2 & 0.12 & 1.20 & 35 & 0.32 & TRUE & 0.18 & 0.01 \\
 3 & 0.02 & 1.20 & 40 & 0.28 & TRUE & 0.18 & 0.01 \\[1mm]
\arrayrulecolor{gray}\hline
\end{tabular}
\caption{Diagnostic on various data sets with varying smoothness. $\ell$ refers to the minimum valid length-scale that can be accurately estimated determined by $m$ and $c$, except for the first iteration which is the initial guess of the length-scale. RMSE* refers to the root mean squared error computed with the exact GP model as the reference.}
  \label{tab_caseI}
\end{table}

Figure~\ref{fig9_Posteriors_exI} shows the posteriors predictive distributions of the exact GP and the HSGP models, the later using the parameter values as the third iterative step, $c=1.2$ ($L=c\cdot 1= 1.2$; see eq.~\eqref{eq_boundary}) and $m=40$ basis functions. In order to do model comparison a spline-based model with 60 knots is also fitted and plotted jointly with the GP and HSGP models. The spline-based model is fitted using the thin plate regression spline approach by \citet{wood2003thin} and implemented in the R-package \textit{mgcv} \citep{wood2011mgcv}. A Bayesian approach is used to fit this spline model using the R-package \textit{brms} \citep{burkner2017brms}. The sample observations are plotted as circles and the out-of-sample or test data are plotted as crosses. The posteriors of the three models, exact GP, HSGP and spline, are similar in the interpolation regions of the input space. However, when extrapolating the spline model solution clearly differs from the exact GP and HSGP models as well as the actual observations.

\begin{figure}
\centering
\includegraphics[width=0.5\textwidth]{fig9_Posteriors_exI.png}
\caption{Posterior predictive means of the proposed HSGP model, the exact GP model, and the spline model. 95\% credible intervals are plotted as dashed lines.}
  \label{fig9_Posteriors_exI}
\end{figure}
